\chapter{Publications}

The work presented in this thesis is based on the following publications:

\vspace{0.5cm}

\begin{description} [labelindent=1cm,style=multiline,leftmargin=3cm]
	\item [Paper 1:] \textbf{Christian Michelsen}\textsuperscript{\textdagger}, Mikkel W. Pedersen\textsuperscript{\textdagger}, Antonio Fernandez-Guerra, Lei Zhao, Troels C. Petersen, Thorfinn S. Korneliussen (2022). ``metaDMG: An Ancient DNA Damage Toolkit''. Submitted to Methods in Ecology and Evolution. \\
	\item [Paper 2:] \textbf{Christian Michelsen}\textsuperscript{\textdagger}, Christoffer C. Jørgensen\textsuperscript{\textdagger}, Mathias Heltberg, Mogens H. Jensen, Alessandra Lucchetti, Pelle B. Petersen, Troels C. Petersen, Henrik Kehlet (2022). ``Preoperative prediction of medical morbidity after fast-track hip and knee arthroplasty -- a machine learning based approach''. In review at BMJ Open. \\
	\item [Paper 3:] Mathias S. Heltberg\textsuperscript{\textdagger}, \textbf{Christian Michelsen}\textsuperscript{\textdagger}, Emil S. Martiny, Lasse E. Christensen, Mogens H. Jensen, Tariq Halasa and Troels C. Petersen (2022). ``Spatial Heterogeneity Affects Predictions from Early-Curve Fitting of Pandemic Outbreaks: A Case Study Using Population Data from Denmark''. Published in: Royal Society Open Science 9.9. issn: 2054-5703. doi: 10.1098/rsos.220018.  \\
	\item [Paper 4:] Susmita Sridar\textsuperscript{\textdagger}, Mathias S. Heltberg\textsuperscript{\textdagger}, \textbf{Christian Michelsen}\textsuperscript{\textdagger}, Judith M. Hattab, Angela Taddei (2022). ``Microscopic single molecule dynamics suggest underlying physical properties of the silencing foci''. Unpublished paper draft.
\end{description}


\vspace{0.5cm}
\noindent Shared first authorship is indicated with a dagger (\textdagger) next to the name. \\

\noindent The appendix contains two papers of which I am a co-author, see \autoref{appendix:kapk} and \autoref{appendix:anaemia}. The appendix further contains two reports published by Statens Serums Institut that are based on my research during my Ph.D., see \autoref{appendix:ssi-report} and \autoref{appendix:ssi-notat}. These appendices are further explained in \autoref{chapter:introduction}.


% \begin{refsection}
%     % \settoggle{bbx:articletitle}{true}%
%     \nocite{heltbergSpatialHeterogeneityAffects2022a}
%     % \nocite{refB}
%     \defbibnote{lop_prescript}{
%         The work presented in this thesis is based on the following publications
%     }
%     \defbibnote{lop_postscript}{
%         Some publications are currently under peer review
%     }
%     \printbibliography[
%         segment=\therefsegment,
%         heading=bibintoc,
%         title={List of Publications},
%         prenote=lop_prescript,
%         postnote=lop_postscript
%     ]
% \end{refsection}